On the Synchronization Power of the Openflow Data Plane

### 2015-03-17 old-inband-STN.tex
stefan's changes

%[[PK sorry for stubborness :)
\title{On the Synchronization Power of Openflow Data Plane} 
%]]

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
In this paper, we show that the standard Openflow
protocol is powerful enough to implement a powerful synchronization
abstractions exporting \emph{transactional} accesses to the components
of the data-plane configuration, with \emph{all-or-nothing} semantics. 
The abstraction can be implemented \emph{in band}, without any
additional synchronization between the controllers.
We describe how to leverage the transactional abstraction to
atomically resolve conflicts between concurrent policy updates
\emph{on-the-fly}, without affecting the ongoing traffic.
%[[PK
%In particular, we describe the implementation of test\&set primitives
%as well as multi-writer consensus objects which can be used
%to implement atomic transactions with all-or-nothing semantics.
%Interestingly, our approach allows controllers to detect and
%resolve very general notions of conflicts, not only, e.g., conflicts
%due to overlapping flow spaces but also conflicts based on network load.
%]]
\end{abstract}

@@

%\textbf{}
\paragraph{Our Contributions}
In this paper, we show that
Openflow~\cite{of-spec}---the standard SDN protocol today---can
be used to implement powerful, high-level
synchronization objects \emph{in-band}.
In particular, we define a transactional synchronization abstraction
that exportes \emph{atomic} (\emph{all-or-nothing} access to multiple
components of an SDN switch configuration, similarly to the way
transactional memory~\cite{stm-st95,tm-book} provides atomic access to
multiple memory locations.
To implement this synchronization abstraction, we rely on several
standard OpenFlow features, including \emph{message bundling},
\emph{check-overlap} flag, and \emph{grouping}~\cite{of-spec}.

%The container object is reminiscent of the hypothetical multi-write
%shared-memory model which is known to enable solutions for important
%synchronization problems, such as consensus, impossible to solve
%otherwise~\cite{multi-objects}.
%Indeed, we show how to exploit the container object to implement
%consensus among controllers.
We show how the transactional abstraction helps consistency of network
management on the control plane, while preserving consistency of the
traffic processed on the data plane, and we show how to implement
\emph{packet-atomic} rule updates~\cite{network-update}: even though
an update .
% and more complex
%\emph{atomic transactions}.
We also sketch how to implement the \emph{read-modify-write} primitives
and \emph{control-plane} synchronization, assumed voluntarily to exist
by the recent proposal of a
consistent composition system for forwarding policies~\cite{cpc}.

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  Openflow
protocol and discuss its features which are relevant for our work.
In Section~\ref{sec:vision}, we introduce our vision and the container consensus object 
and present its implementation in OpenFlow.
We then discuss applications (Section~\ref{sec:apps}) as
well as extensions (Section~\ref{sec:discussion}).
We review related literature in Section~\ref{sec:relwork}, and conclude
in Section~\ref{sec:conclusion}.

@@

\section{Openflow Background}\label{sec:background}

@@

In a nutshell,
Openflow
is based on a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules (flow entries). Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline; during the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information. Concretely, instructions can be
used to define additional tables to be visited (goto instruction),
to modify the set of to-be-applied actions (either by appending,
deleting, or modifying actions), or immediately apply some
actions to the packet. A meta-data field can be used to exchange
information between tables; it is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

The Openflow standard (as of \textbf{TODO}) offers four interesting
features which we will exploit to implement the transactional memory abstraction:

@@

\item \textbf{Conditional Application:}
The Openflow standard also supports the \emph{conditional} creation of
flow entries. A specific $\textsf{OFPFF\_CHECK\_OVERLAP}$ flag
set in the flow creation message ensures
that the flow entry is created if and only if it does not \emph{conflict}
with any other flow entry currently in the flow table.
This  conflict detection mechanism, combined with bundling,
allows us to implement specific types of \emph{transactions} that may
atomically install multiple rules if certain compatibility conditions
are met.

@@

%\section{The Container Object}\label{sec:vision}
\section{Transactional synchronization}\label{sec:vision}

@@

We assume that every transaction has a unique identifier. 
Every operation within a transaction generates a number of  OpenFlow
messages. Each of the messages can be wrapped 
is marked as a 
in a \emph{bundle} message
%,where the transaction's identifier is used as the bundle id. 
.
Each bundle message carries a bundle id, where messages with same bundle id are considered to belong to same bundle and are performed atomically.

%The OpenFlow messages that we use (besides )

@@

\section{Applications}\label{sec:apps}

@@

\subsection{Policy updates}

@@

\section{Discussion and Extensions}\label{sec:discussion}

FIXME: we could however even do stronger stuff: collision is only exact, not using masks: do not use the ternary nature

\paragraph{Beyond Container Objects}
FIXME: we can even do more general things!

\paragraph{Older Openflow Versions} FIXME: describe matrix approach

@@

@@

\textbf{PK: anything about your ``brain'' paper, feel related?}

@@

@@

Here we describe a second implementation of the Compare\&Set abstraction that is based on entirely on flow entry creations and deletions (FlowMod commands).

>>>>

%[[PK sorry for stubborness :)
\title{On the Synchronization Power of the Openflow Data Plane}
%\\ {\Large Towards Atomic Transactions In-Band}}
%]]

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence 
of operations which is performed in an atomic, serializable manner, providing all-or-nothing semantics. 
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol. 
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}. 
In particular, we present a
transactional interface which supports arbitrary atomic
transactions 
over the switch configuration space: Our transactions 
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and controller interactions. We discuss different use cases
for our interface (consistent policy installation, load-balancing, consensus),
and also describe how to 
perform basic forms of policy compositions in-band.
\end{abstract}

@@

The distributed computing literature offers 
a wide range of abstractions 
and techniques to solve distributed synchronization and consistency problems. 
A particularly interesting concept in distributed computing is a \emph{transactional memory}: 
a shared memory which supports atomic transactions, describing a sequence of operations 
to be executed with an all-or-nothing semantics. Transactional memories can be used to
 solve many 
classic distributed computing problems, such as consensus.

This paper investigates the design of a transactional interface for software-defined
networks. In particular, we argue that the switch configuration state can be seen
as a shared memory which is modified concurrently by multiple controllers.
Being able to provide more complex and atomic sequences of transactions over this configuration
space can be appealing to resolve synchronization problems.

\paragraph{Our Contributions} 
This paper introduces a very powerful transactional interface
 for software-defined networks. This interface is implemented
 in-band, and
 using Openflow~\cite{of-spec} (the standard SDN protocol today) features only.
Indeed, we find that Openflow's \emph{bundle} or \emph{rule overlap check}
features can serve as synchronization primitives. 
However, without further mechanisms, their use is limited:
with a bundle alone it is not possible to react to
and change switch configuration state in any non-trivial way,
and with the rule overlap check it is not possible to react to
any non-trivial update conflicts.

In contrast, the transactional interface presented in this paper supports very general atomic
transactions over the switch configuration space, similar to classic transactional memory~\cite{stm-st95,tm-book}. 
For instance, a transaction allows to modify switch configurations in an interactive 
and computational fashion, as well as to define complex notions
of conflicts. 
Our implementation is based on distributed identifiers and makes use of two 
new synchronization objects for managing these identifiers:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the 
currently used identifier.
Thus, our work also provides the missing link for the read-modify-write object
postulated in~\cite{cpc}.
Our work also contributes to the ongoing discussion of what can be implemented
in-band in Openflow~\cite{compute,reclaim}, by taking a first look
at the relevant problem of distributed
synchronization.

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  Openflow
protocol and discuss its features which are relevant for our work.
%In Section~\ref{sec:vision}, we introduce our vision and the container consensus object
%and present its implementation in OpenFlow.
%We then discuss applications (Section~\ref{sec:apps}) as
%well as extensions (Section~\ref{sec:discussion}).
FIXME
We review related literature in Section~\ref{sec:relwork}, and conclude
in Section~\ref{sec:conclusion}.

@@

\section{Background and Motivation}\label{sec:background}

@@

\textbf{Background.} In a nutshell,
Openflow
is based on a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules (flow entries). Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline; during the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information. Concretely, instructions can be
used to define additional tables to be visited (goto instruction),
to modify the set of to-be-applied actions (either by appending,
deleting, or modifying actions), or immediately apply some
actions to the packet. A meta-data field can be used to exchange
information between tables; it is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

\textbf{Useful features for synchronization.} The Openflow standard (as of \textbf{TODO}) offers four interesting
features which we will exploit to implement the transactional memory abstraction:

@@

\item \textbf{Check Overlap:}
The Openflow standard also supports the \emph{conditional} creation of
flow entries. A specific $\textsf{OFPFF\_CHECK\_OVERLAP}$ flag
set in the flow creation message ensures
that the flow entry is created if and only if it does not \emph{conflict}
with any other flow entry currently in the flow table.
This  conflict detection mechanism, combined with bundling,
allows us to implement specific types of \emph{transactions} that may
atomically install multiple rules if certain compatibility conditions
are met.

@@

\textbf{Limitations.}  While these features are useful primitives, 
building complex transactions from these features
is a non-trivial task, as we will show in the following. 
A bundle can be understood as a transaction:
 it defines a sequence of read and write operations 
 on the switch state which are executed atomically.
 However, such bundle transactions come with many 
 limitations. A bundle transaction does not allow
 to change the switch state in any non-trivial way,
 as the read and write actions in the bundle do 
 not support any conditional modifications or computations which 
 depend on the read results. By leveraging
 the check overlap feature, it is at least possible
 to abort transactions depending on certain conditions
 in the overlapping
 flow spaces. However, in many cases, for example
 a load-balancing application,
 it can be interesting to define more general
 dependencies, which depend on, from a logical perspective,
 seemingly
 completely independent flow spaces, 
 which however share the underlying physical network.
In other words, the bundle feature only allows us to 
define ``one-shot transactions'',
transactions which do not support any further controller interactions. 
Given
the limited computational options the Openflow interface of a switch offers, 
it is difficult to implement more complex behavior.

%\section{The Container Object}\label{sec:vision}
\section{Transactional Synchronization}\label{sec:vision}

This section describes how to introduce more powerful transactions. We first present a new synchronization object to

@@

We assume that every transaction has a unique identifier.
Every operation within a transaction generates a number of  OpenFlow
messages using \textsc{FlowMod} commands.
We therefore assume that each of these OpenFlow messages is marked as  a \emph{bundle} message,
where the transaction's identifier is used as the bundle id.

@@

\section{Discussion of Use-Cases}\label{sec:apps}

@@

FIXME: load-balancing, consensus, etc.

@@

@@

We conclude our technical contribution by showing that sometimes,
powerful synchronization mechanisms can be implemented

 This section show how to implement

@@

@@

Our work also contributes to the ongoing discussion of what can be implemented
in-band in Openflow~\cite{compute,reclaim}, by assuming the relevant perspective of distributed
synchronization.

@@

This section describes an implementation without group tables.
